---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
The file "activity.zip" was originally from [this url](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).

## Loading and preprocessing the data
We start by unzipping the file and loading the csv file within.
```{r}
unzip("activity.zip")
act_raw <- read.csv("activity.csv")
```
We then take a look on what's inside:
```{r}
str(act_raw)
```
From this we realize that the dates are factors. Not that bad, but the best  
thing to do is to convert them to actual dates. We also transform the  
intervals into factors.
```{r}
act_pro <- transform(act_raw, interval = factor(interval),
                     date = as.Date(date, "%Y-%m-%d"))
```
Let's check this out:
```{r}
str(act_pro)
```

## What is mean total number of steps taken per day?
For this, we will need additionnal packages:
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
```
We start by grouping the data by date, we will do this by creating a new  
dataframe:
```{r}
df_by_day <- act_pro %>% group_by(date) %>% 
    summarise(steps_by_day = sum(steps, na.rm = F),
              nb_NA = sum(is.na(steps)))
table(df_by_day$nb_NA)
```
The variable "nb_NA" clearly shows the repartition of the NA's : Either a  
day has 288 values (one by interval) or none (288 NA's). This will help us  
in the future.  
  
For now, we want a histogram of the steps taken each day. This will show the repartition of the sums.
```{r}
hist1 <- qplot(steps_by_day, data = df_by_day, binwidth = 1500)+
    ylab("Count (number of days)")+
    xlab("Number of steps (unit)")+
    labs(title = "Histogram of the total steps taken by day.",
         subtitle = "Median in red.")+
    geom_vline(xintercept = median(df_by_day$steps_by_day, na.rm = T),
               color = 2)
print(hist1)
```
  
We can clearly see that the totals gravitate around the median (see below)  
with a rather central pattern.
```{r}
summary(df_by_day$steps_by_day)
```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
